Disentangling patterns of gene expression in high grade serous ovarian cancer by matrix factorization

### 2020-08-11 MSc_Dissertation_B156476.tex
Added integrative tables for main results and for combined ('BOTH') experiments; discussion of integrative results; other miprovement; some figures removed

\tableofcontents

\newpage

@@

\begin{enumerate}
\item
Identification of a consistent gene set between the TCGA and AOCS datasets.
\item
Unsupervised metagene extraction by matrix factorization methods : NMF, ICA and PCA.
\item
Metagene selection for robustness by k-means clustering of bootstrap resampled factorizations evaluated by silhouette score, based on the TCGA dataset.
\item
Investigation of biological significance of determined factors by gene enrichment analysis against the Gene Ontology (GO).
\item
Computation by least squares optimisation of metasamples associated with with each TCGA derived metagene against TCGA and AOCS datasets.
\item
Alignment of available metadata with computed metasamples, for TCGA and AOCS.
\item 
Survival analysis on TCGA and AOCS to investigate relationship between metasamples and patient survival.
\item
Scatter plots and heat maps to investigate the relationships between metasamples and genomic features available for the AOCS dataset.
\end{enumerate}

@@

\subsubsection{Tools}
\begin{itemize}
\item Python 3.6
\item PyCharm Integrated Development Environment
\item Numpy for high performance matrix manipulation
\item Pandas for data frame handing
\item Matplotlib and Seaborn for general plotting
\item Scikit-learn for matrix factorization and k-means clustering
\item GOATOOLS package for gene enrichment analysis against GO
\item Lifelines package for survival analysis.
\end{itemize}

A full list of software and libraries used, with versions, is given in appendix \ref{sec-software-versions}.

\subsection{Gene set intersection}
\label{sec:gene-set-intersection}

@@

\subsection{Metagene selection by cluster coherence}
\label{sec-metagene-selection}
Deciding on the number of components (factors, metagenes) to extract -- that is the factorization rank -- is key.  Taking more components results in more accurate representation of the observed expression matrix and provides more avenues to explore the underlying biology.  However, it is important that the metagenes are \emph{stable}, that is that they have reliable meaning when transfered to other datasets.   There are two sources of variation or instability to consider.

@@

\subsection{Investigating correlation between metasamples and genomic features}

@@

@@

The described methodology was implemented in Python.
%except for the KEGG gene enrichment analysis which was implemented in R.
All code is available in a github repository: \url{https://github.com/ipoole/HgsocTromics}.  Good software engineering practice has been followed, with an object-oriented design, frequent commits and unit testing.  Unit tests are based on tiny expression matrices, just 100 genes by 10 patients, thus the whole test suite of over 60 tests executes in around 40 seconds and can be run frequently.  These tests are particularly valuable when re-factoring code.  The total codebase is approximately 2,900 lines of Python.

All plots are generated in vector graphics pdf format to ensure smooth scaling to any resolution.  A consistent colour scheme of \textcolor{blue}{blue}, \textcolor{orange}{orange} and \textcolor{olive}{green} was used for plots relating to NMF, ICA and PCA respectively.  Metagenes are consistently referred to by, for example, ``NMF-2-of-3" -- meaning the 2nd component of the rank $k=3$ NMF factorization.

@@

For this reason it is appropriate to perform all metagene extraction on the larger TCGA dataset, in the expectation that the obtained metagenes will be more robust and likely to better generalise to other datasets.

@@

\subsection{Gene sets identified by metagenes of different methods intersect in some cases}

@@

Considering the biological significance of each of the 11 extracted metagenes, with reference to the GO enrichment results:
\begin{description}
%
\item[NMF 1-of-3]: this is related to the cellular structures and processes of the extra-cellular matrix (ECM), that is the proteins such as collagens which mediate the three-dimensional organisation of cells in a tissues.  ECM molecular composition will vary substantially between tissue types, but is also known to play a part in many disease processes \cite{Theocharis2019}.  Thus, this metagene may simply reflect heterogeneity of tissues in the biopsied sample, or may have some deeper disease related significance.
%
\item[NMF 2-of-3]: here we see enrichment of genes relating to the ribosomal subunit and the processes of RNA binding, implying perhaps a link with assembly of the ribosomal RNA-protein complex.   Ribosomes are known to have a role in carcinogenesis, by dysregulation the RNA $\rightarrow$ protein translation, or mutations in ribosomal subunits impacting on cellular metabolism \cite{Sulima2017}.
%
\item[NMF 3-of-3, ICA 1-of-5 and ICA 2-of-5]: no significant biological enrichment found.
%
\item[ICA 3-of-5]: this component relates to processes of multicellular / extracellular organisation, and the ECM.  It is therefore similar to NMF 1-of-3 above.
\item[ICA 4-of-5]: this component is seen to relate to the ribosomal subunit (as NMF 2-of-3), additionally membrane proteins in the respiratory complex, mitochondrial and the NADH dehydrogenase complex.  Processes of organonitrogen biosynthesis are also highlighted. 
%
\item[ICA 5-of-5]: processes relating to regulation of immune response are enriched in this component, featuring genes from the major histocompatibility complex group, in particular the HLA- genes, allowing the immune system to recognise self from non-self.  GO terms relating to the endoplasmic reticulum (ER) are also highlighted.  It may be that this component is mainly sensitive to the immunohistochemical signature of patients, and therefore not of clinical interest.
%
\item[PCA 1-of-3]: there is mention of ECM related terms, as for ICA 3-of-5 and NMF 1-of-3.   However, there are also terms relating to the regulation of angiogenesis.  It is known that tumours have a need for increased blood supply, and that expression level of factors promoting angigenesis are associated with aggresiveness of tumour growth \cite{Nishida2006}.  
We have already seen (section \ref{sec-subtype-classification}, \cite{Glass2015}) there is evidence for distinct subtypes of OC distinguished by angiogenesis related genes, and that these subtypes have been found to inform clinical outcome \cite{Glass2015}.  Thus, this component might contain useful prognostic value.
%
\item[PCA 2-of-3]: no significant biological enrichment found.
%
\item[PCA 3-of-3]: this component has some similarity with ICA 5-of-5, in that it refers to regulation of immune processes (HLA- genes) and ER membrane.   However, chemotaxis (cell movement) is also highlighted.
\end{description}

We have noted above that both NMF 1-of-3 and ICA 3-of-5 may be related to tissue type heterogeneity.  We have also seen from metasample heatmap analysis that these two metagenes are associated with closely correlated metasamples (in AOCS), and also to have a moderately high Jaccard similarity of 0.39.  Further, we observed that these components correlate negatively with cellularity.  This is as one would expect: low cellularity implies a mixture of tumour and non-tumour cell types, resulting in variation of ECM composition.

Interestingly, the above summary of the gene enrichment results tentatively suggests component PCA 1-of-3 is most likely to have clinical prognostic value, based on a cursory look at the literature.  This perhaps supports an earlier observation that, while not conclusive, component PCA-1-of-3 comes closest to demonstrating correlation with patient survival (section \ref{sec-survival-analysis}, figure \ref{fig-summary-survival-plots}).

How should we interpret those four components which demonstrate no significant enrichment with biological meaning?  It is notable that all three methods have at least one such component.  Among these four, PCA 1-of-3 and ICA 1-of-5 have high (the highest) Jaccard similarity of 0.61.   For all other pairs among these four Jaccard similarity is negligible.  Perhaps these two components are picking up on similar technical variation between the samples?

@@

@@

\subsection{Metasample correlation with genomic features}

From the grid of scatter plots shown in figure \ref{fig-genomic-scatters} it can be seen there are some significant correlations.  
The sense (sign) of the correlation is arbitrary -- particularly for ICA and PCA -- recall normalisation w.r.t $180^\circ$ was applied.

The significant correlations are summarised in table \ref{tab-genomic-scatter-summary}.  
WGD and Cellularity show correlation with at least one component from all three methods and indeed three components correlating with WGD also correlate with Cellularity and in the same sense.  
This might suggest simply that WGD and Cellularity are themselves correlated, but this is not so: $r=-0.14, p=0.201$ for Point-Biserial correlation (data not shown).  % Latent factor??
HR Detect correlates with two NMF and one ICA component, not intersecting with those above, suggesting distinct biological processes are involved.

The heatmap dendrogram (figure \ref{fig-heatmap-TCGA-AOCS}) shows close correlation between metasamples pairs (NMF-1-of-3, ICA-5-of-5), (ICA-1-of-5, PCA-2-of-3) and (NMF-1-of-3, ICA-3-of-5).

\begin{figure}[htb!]
\begin{center}
\includegraphics[scale=0.7, draft=false]{\plotspath{HeatMaps/clustered_heatmap_TCGA_AOCS.pdf}} 
\caption{Heatmap of the metasamples matrix for the AOCS dataset, computed from metagenes factorized from the TCGA dataset.   Several patient metadata variables are shown in columns to the left.}
\label{fig-heatmap-TCGA-AOCS}
\end{center}
\end{figure}

\begin{table}[htb!]
\begin{center}
\begin{tabular}{|l|l|}
\hline
Genomic feature & Significant correlations \\
\hline 
WGD & NMF-1-of-3, ICA-5-of-5, PCA-3-of-3 \\
Cellularity & NMF-1-of-3, ICA-3-of-5, ICA-5-of-5, PCA-3-of-3 \\
HR Detect & NMF-2-of-3, NMF-3-of-3, ICA-2-of-5 \\
Mutational load & None \\
CNV load & None \\
SV load & NMF-2-of-3 \\
\hline
\end{tabular}
\end{center}
\caption{Summary of metasample components showing a significant correlation with each genomic feature.}
\label{tab-genomic-scatter-summary}
\end{table}

\section{Matrix factorization discovers and effectively removes batch effects}

Here are presented results of applying the three MF methods to the combined AOCS and TCGA datasets.

@@

@@

This work has thrown up several possible lines of further study:
\begin{enumerate}
\item
{\bf Algorithm hyper-parameters}.  NMF and ICA algorithms have a number of hyper-parameters, only a few of which were explored in this work.  For example, NMF has parameters to encourage sparsity through L1 regularization.   This should result in many (most?) metagene elements reducing to zero, and might offer a more robust means of selecting candidate genes to feed into gene enrichment analysis -- replacing the current arbitrary 3 SD from the mean rule.
\item
{\bf Discovering batch effects}.  It has been claimed (e.g. \cite{Stein-OBrien2018}) that matrix factorization methods are an effective means of identifying and removing batch effects.  This could be explored by horizontally concatenating the TCGA and AOCS datasets (after gene set intersection), producing an $n=80+274=354$ dataset with a substantial batch artefact.  Factorizations of this combined dataset should be more robust because of the larger $n$.  Does each of NMF, ICA and PCA naturally produce a component which correlates with batch?
\item
{\bf Selecting components from several factorization ranks}.
In the current work, a single rank was selected for each method -- 3, 5 an 3 for NMF, ICA and PCA respectively.  Unlike PCA, components extracted by NMF and ICA  do not `nest' with rank; that is adding rank in general yields a new set of components.  Thus one might, for NMF and ICA, perform factorizations at $k=2,3,4,5$, say, yielding $2 \times 15 = 30$ potential components in total.  Jaccard similarity could be used to identify that subset of components which had the least overlap in detected genes.   In this way, a larger number of components could be obtained without use of high factorization ranks which we have seen to be unstable.
\item
{\bf Cross-dataset factorization stability}.  We went to considerable effort -- through bootstrap sampling and cluster analysis -- to select factorization ranks which we hoped would be robust and generalise well to other datasets.   A way to confirm this robustness would be to perform the factorization / clustering pipeline on \emph{both} datasets separately, yielding two sets of metagenes.  In a perfect world, these metagenes would pair up identically (in-so-far as the two datasets had identical technical characteristics, drawn from identical population of patients).  The Jaccard similarity heatmap could be used to verify this, and reject components which showed low cross-dataset similarity.
\item
{\bf Contrasting gene expression patterns between cancers}.  The current work set out to find patters of gene expression in HGSOC.  Yet it is hard to say the highlighted patterns are specific to HGSOC, or generic to all cancers.  One way of teasing out cancer specific patterns might be to take a similar approach to the previous item, but instead of looking for metagenes which are consistent between datasets, find those which are distinct.  That said, there will likely be many more sophisticated approached in the literature, given the clinical importance of the topic.
\item
{\bf Systematic comparison with published gene expression patterns}.  In analysing the meaning of each metagene above, some tentative links were made with the research literature.   But this was anecdotal and frankly not particularly scientific; searching on PubMed for GO terms associated with HGSOC and looking at one or two hits!  A more systematic and ideally automated approach is required.  The \href{https://www.ncbi.nlm.nih.gov/geoprofiles}{Geo Profiles database} at NCBI, or the \href{https://www.ebi.ac.uk/gxa/home}{Expression Atlas} at EMBL-BI might be possible starting points.

@@

\begin{figure}[htb!]
\begin{center}
\includegraphics[scale=0.6]{\plotspath{BoxAndScatters/wgd_boxplot_NMF_TCGA_AOCS.pdf}}
\includegraphics[scale=0.6]{\plotspath{BoxAndScatters/wgd_boxplot_ICA_TCGA_AOCS.pdf}}
\includegraphics[scale=0.6]{\plotspath{BoxAndScatters/wgd_boxplot_PCA_TCGA_AOCS.pdf}}
\caption{Boxplots showing relationship between metasamples (derived from TCGA metagenes, applied to AOCS data) and whole genome doubling (WGD) status.  Metasamples from NMF, ICA and PCA are shown separately.  To allow convenient visualisation, each metasample is normalised to zero mean unit variance, hence between metasample variation is removed.}
\end{center}
\label{fig-wgd-boxplots}
\end{figure}

@@

\begin{figure}[htb!]
\begin{center}
\includegraphics[scale=0.5]{\plotspath{SurvivalAnalysis/multiple_kaplan_meier_os_BOTH_BOTH.pdf}} 
\caption{Kaplan-Meier plots for each metasample component, stratified at the median value, for the combined AOCS + TCGA dataset}
\label{fig-kaplan-meier-os-BOTH}
\end{center}
\end{figure}

>>>>

\setlength{\parskip}{0em}
\tableofcontents
\setlength{\parskip}{1em}
\newpage

@@

\begin{enumerate}
\item
Identification of a consistent gene set between the TCGA and AOCS datasets.
\item
Unsupervised metagene extraction by matrix factorization methods : NMF, ICA and PCA.
\item
Metagene selection for robustness by k-means clustering of bootstrap resampled factorizations evaluated by silhouette score, based on the TCGA dataset.
\item
Investigation of biological significance of determined factors by gene enrichment analysis against the Gene Ontology (GO).
\item
Computation by least squares optimisation of metasamples associated with with each TCGA derived metagene against TCGA and AOCS datasets.
\item
Alignment of available metadata with computed metasamples, for TCGA and AOCS.
\item 
Survival analysis on TCGA and AOCS to investigate relationship between metasamples and patient survival.
\item
Scatter plots and heat maps to investigate the relationships between metasamples and genomic features available for the AOCS dataset.
\item
Investigation of batch effects by repeating of metagene extraction and heatmaps on the \emph{combined} AOCS and TCGA datasets.
\end{enumerate}

@@

\subsubsection{Tools}
\begin{itemize}
\itemsep -0.5em 
\item Python 3.6
\item PyCharm Integrated Development Environment
\item Numpy for high performance matrix manipulation
\item Pandas for data frame handing
\item Matplotlib and Seaborn for general plotting
\item Scikit-learn for matrix factorization and k-means clustering
\item GOATOOLS package for gene enrichment analysis against GO
\item Lifelines package for survival analysis.
\end{itemize}
A full list of software and libraries used, with versions, is given in appendix \ref{sec-software-versions}.

\subsection{Syncrhonisation of gene sets accross AOCS and TCGA}
\label{sec:gene-set-intersection}

@@

\subsection{Metagene extraction by cluster coherence}
\label{sec-metagene-selection}
Deciding on the number of components (factors, metagenes) to extract -- that is the factorization rank -- is key.  Taking more components results in more accurate representation of the observed expression matrix and provides more avenues to explore the underlying biology.  However, it is important that the metagenes are \emph{stable}, that is that they have reliable meaning when transfered to other datasets.   There are two sources of variation or instability to consider.

@@

\subsection{Investigating correlation between metasamples and genomic features}
\label{sec-genomic-features}

@@

\subsection{Investigating the impact of batch effects by analysis of a combined dataset}

It has been claimed (e.g. \cite{Stein-OBrien2018}) that matrix factorization methods are an effective means of identifying and removing batch effects.  To verify this, the TCGA and AOCS datasets (after gene set intersection), were horizontally concatenated producing an $n=80+274=354$ patient dataset with a substantial batch artefact.  Survival data is available for both datasets, so this was preserved as metadata associated with the combined datasets. The previously described analyses of metagene extraction, metasample computation and clustered heatmap and survival analysis were repeated on this dataset.  Ideally, the factorization rank selected for each method should be reconsidered, since the larger dataset could justify extracting more components.  For simplicity this was not done however, the previously selected $k = 3,5,3$ for NMF, ICA and PCA respectively being maintained.   Note however that we should not expect the extracted metagenes and derived metasamples to be equivalent. Genomic feature correlation w.r.t these newly extracted components was performed using the AOCS datasets.  Theses results were inspected; is there evidence of some metasamples aligning with technical batch effects while others align with biological processes?

@@

The described methodology was implemented in Python.
%except for the KEGG gene enrichment analysis which was implemented in R.
All code is available in a github repository: \url{https://github.com/ipoole/HgsocTromics}.  Good software engineering practice has been followed, with an object-oriented design, frequent commits and unit testing.  A base class is used to present a uniform interface for the three factorization methods.  Intermediate results are cached with lazy computation for efficient working without ordering dependency.  Unit tests are based on tiny expression matrices, just 100 genes by 10 patients, thus the whole test suite of over 65 tests executes in around 45 seconds and so can be run frequently.  The total codebase is approximately 2,900 lines of Python.

All plots are generated in vector graphics pdf format to ensure smooth scaling to any resolution.  A consistent colour scheme of \textcolor{blue}{blue}, \textcolor{orange}{orange} and \textcolor{olive}{green} is used for plots relating to NMF, ICA and PCA respectively.  Metagenes are consistently referred to by, for example, ``NMF-2-of-3" -- meaning the 2nd component of the rank $k=3$ NMF factorization.

@@

For this reason it is was decided to perform all metagene extraction on the larger TCGA dataset, in the expectation that the obtained metagenes will be more robust and likely to better generalise to other datasets.

@@

\subsection{Jaccard similarity shows metagenes identify intersecting gene sets in some cases}

@@

Considering the biological significance of each of the 11 extracted metagenes, with reference to the GO enrichment results:
\begin{description}
%
\item[NMF 1-of-3]: this is related to the cellular structures and processes of the extra-cellular matrix (ECM), that is the proteins such as collagens which mediate the three-dimensional organisation of cells in a tissues.  ECM molecular composition will vary substantially between tissue types, but is also known to play a part in many disease processes \cite{Theocharis2019}.  Thus, this metagene may simply reflect heterogeneity of tissues in the biopsied sample, or may have some deeper disease related significance.
%
\item[NMF 2-of-3]: here we see enrichment of genes relating to the ribosomal subunit and the processes of RNA binding, implying perhaps a link with assembly of the ribosomal RNA-protein complex.   Ribosomes are known to have a role in carcinogenesis, by dysregulation the RNA $\rightarrow$ protein translation, or mutations in ribosomal subunits impacting on cellular metabolism \cite{Sulima2017}.
%
\item[NMF 3-of-3, ICA 1-of-5 and ICA 2-of-5]: no significant biological enrichment found.
%
\item[ICA 3-of-5]: this component relates to processes of multicellular / extracellular organisation, and the ECM.  It is therefore similar to NMF 1-of-3 above.
\item[ICA 4-of-5]: this component is seen to relate to the ribosomal subunit (as NMF 2-of-3), additionally membrane proteins in the respiratory complex, mitochondrial and the NADH dehydrogenase complex.  Processes of organonitrogen biosynthesis are also highlighted. 
%
\item[ICA 5-of-5]: processes relating to regulation of immune response are enriched in this component, featuring genes from the major histocompatibility complex (MHC) group, in particular the HLA- genes, allowing the immune system to recognise self from non-self.  GO terms relating to the endoplasmic reticulum (ER) are also highlighted.  It may be that this component is mainly sensitive to the immunohistochemical signature of patients, and therefore not of clinical interest.
%
\item[PCA 1-of-3]: there is mention of ECM related terms, as for ICA 3-of-5 and NMF 1-of-3.   However, there are also terms relating to the regulation of angiogenesis.  It is known that tumours have a need for increased blood supply, and that expression level of factors promoting angigenesis are associated with aggresiveness of tumour growth \cite{Nishida2006}.  
We have already seen (section \ref{sec-subtype-classification}, \cite{Glass2015}) there is evidence for distinct subtypes of OC distinguished by angiogenesis related genes, and that these subtypes have been found to inform clinical outcome \cite{Glass2015}.  Thus, this component might contain useful prognostic value.
%
\item[PCA 2-of-3]: no significant biological enrichment found.
%
\item[PCA 3-of-3]: this component has some similarity with ICA 5-of-5, in that it refers to regulation of immune processes (HLA- genes) and ER membrane.   However, chemotaxis (cell movement) is also highlighted.
\end{description}

@@

\subsection{Metasamples correlate with genomic features}

From the grid of scatter plots shown in figure \ref{fig-genomic-scatters} it can be seen there are some significant correlations at play.  These are summarised and analysed in the next section.

@@

\subsection{Integrative analysis of results}

Table \ref{tab-integrative-analysis} summarises the key results of all the forgoing analysis.

\begin{table}[htb!]
\begin{center}
\begingroup
\scriptsize
\renewcommand{\arraystretch}{1.4} % Default value: 1
\begin{tabular}{|l||c|c|c|c|c|c|c|l|l|}
\hline
Metagene & S'val & WGD & Cell. & HRD & Mut.L. & CN.L. & SV.L. & Enriched GO terms & Similar to\\
\hline
%     & S & W & C & H % M & C & S & Enriched % Sim.
NMF-1 & . & Y-& Y-& . & . & . & . & Extra-cellular matrix (ECM) & ICA-3, PCA-1 \\
NMF-2 & . & . & . & Y+& . & . & Y+& Ribo. subunit & - \\
NMF-3 & - & - & - & Y-& . & . & . & . & . \\
\hline
ICA-1 & . & . & . & . & . & . & . & . & PCA-2 \\
ICA-2 & . & . & . & . & . & . & . & . & . \\
ICA-3 & . & . & Y-& . & . & . & Y-& ECM, multicellular org. & NMF-1, PCA-1 \\
ICA-4 & . & . & . & . & . & . & . & Ribo subunit, Mitoch. NADH dehydr. & - \\
ICA-5 & . & Y+& Y+& . & . & . & . & Immune response, MHC & PCA-3 \\
\hline
PCA-1 & Y+& . & Y-& . & . & . & . & ECM, Reg. angiogenesis & NMF-1, ICA-3 \\
PCA-2 & . & . & . & . & . & . & . & . & ICA-1 \\
PCA-3 & . & Y-& Y-& . & . & . & . & Immune response, MHC, chemotaxis & ICA-5 \\
\hline
\end{tabular}
\endgroup
\end{center}
\caption{For each metagenes a 'Y' indicates that a significant correlation was observed with: Survival, WGD, HRD, Mutational load,  CNV load and SV load.  The next column gives a terse summary of the gene enrichment analysis results.  The final column lists the metagenes which are apparently similar -- having a Jaccard similarity of $\geq 0.35$.}
\label{tab-integrative-analysis}
\end{table}
From this summary table we can make some interesting observations:
\begin{description}
\item[ECM relationship to Cellularity?]
We see that NMF-1, ICA-3 and PCA-1 each have an ECM association and we have noted this could be related to tissue type heterogeneity.   These three components (one from each of the methods) are also observed to be correlated with Cellularity.  The sense (+/-1) of the correlation is meaningful for NMF (not so for ICA or PCA) and we see that ECM and thence tissue heterogeneity is negatively correlated with cellularity. This seems reasonable, in that a low Cellularity implies that a greater proportion of surrounding non-tumour tissue is included in the sample, so increasing heterogeneity.   That these three components turn out to be similar by the Jaccard index is as would be expected given their functional similarity.
\item[Angiogenesis link with survival?]
The only component showing (albeit weak) correlation with patient survival is PCA-1.  Interestingly, this same component is also found to be linked with Go terms relating to the regulation of angiogenesis, which in turn is related to aggressiveness of tumour growth and outcomes.  Thus, these observations are telling a consistent story.
\item[Normal cell activity?]
Component ICA-4 highlights the greatest number of biological processes -- indeed the GO lineage map \ref{fig-ICA-4-go-lineage} needed to be pruned to fit on a page.  And yet this component shows no correlation with any disease related factors, neither survival nor genomic features.  It may be that this component is simply picking up on the normal metabolic activity of the cells,
\item[Technical batch effects?]
There are four components (NMF-3, ICA-2, ICA-4 and PCA-2) showing no enrichment of GO terms.  These components also show almost no correlation with disease related factors (the exception being NMF-3 linking with HRD).  With that exception aside, we could conclude that these components have little biological meaning, and are instead picking up technical effects.  I do not know how the TCGA dataset was assembled; perhaps there are multiple batches within the set?
\item[WGD and Cellularity have curious common influences.]
The three components NMF-1, ICA-5 and PCA-3 (one from each method) correlate with both WGD and Cellularity, and in the same sense. 
This might suggest simply that WGD and Cellularity are themselves positively correlated, but this is not so: $r=-0.14, p=0.201$ for Point-Biserial correlation (data not shown).  Neither is there any close similarity between the three components by Jaccard, although ICA-5 and PCA-3 are both linked to immune response and MHC.   Is there some curious non-linear interaction?
\item[Mutational load CNV load are not captured by these components.]  This is somewhat surprising, perhaps the link would emerge with more extracted components.
\end{description}

\subsection{Matrix factorization discovers and effectively removes batch effects}

Here are presented results of applying the three MF methods to the \emph{combined} AOCS and TCGA datasets.

@@

\begin{table}[htb!]
\begin{center}
\begingroup
\scriptsize
\renewcommand{\arraystretch}{1.4} % Default value: 1
\begin{tabular}{|l||c|c|c|c|c|c|c|c|l|l|}
\hline
Metagene & Which & WGD & Cell. & HRD & Mut.L. & CN.L. & SV.L. & Enrichment? \\
\hline
%     & w & W & C & H % M & C & S & Enrichment?
NMF-1 & Y & . & . & . & . & . & . &  Y \\
NMF-2 & Y & Y & Y & . & . &   & . &  Y \\
NMF-3 & . & Y & Y & . & . & . & . &  Y \\
\hline
ICA-1 & Y & . & . & . & . & . & . &  . \\
ICA-2 & Y & . & . & . & . & . & . &  . \\
ICA-3 & . & . & . & . & . &   & . &  Y \\
ICA-4 & . & Y & Y & . & . & . & . &  Y \\
ICA-5 & . & . & . & . & . & . & . &  Y \\
\hline
PCA-1 & . & Y & Y & . & . & . & . &  Y \\
PCA-2 & Y & . & . & . & . & . & . &  . \\
PCA-3 & Y & . & . & . & . & . & . &  . \\
\hline
\end{tabular}
\endgroup
\end{center}
\caption{In a similar format as table \ref{tab-integrative-analysis}, this summarises results of in respect of the \emph{combined AOCS and TCGA} datasets.  The key point here is that a known batch effect has been introduced, reflected in the 'Which' column. Note that since these metagenes have been re-extracted from different data, they are distinct from those analysed in the previous section}
\label{tab-integrative-analysis-BOTH}
\end{table}

@@

This work has thrown up several possible lines of further study:
\begin{enumerate}
\item
{\bf Algorithm hyper-parameters}.  NMF and ICA algorithms have a number of hyper-parameters, only a few of which were explored in this work.  For example, NMF has parameters to encourage sparsity through L1 regularization.   This should result in many (most?) metagene elements reducing to zero, and might offer a more robust means of selecting candidate genes to feed into gene enrichment analysis -- replacing the current arbitrary 3 SD from the mean rule.
\item
{\bf Selecting components from several factorization ranks}.
In the current work, a single rank was selected for each method -- 3, 5 an 3 for NMF, ICA and PCA respectively.  Unlike PCA, components extracted by NMF and ICA  do not `nest' with rank; that is adding rank in general yields a new set of components.  Thus one might, for NMF and ICA, perform factorizations at $k=2,3,4,5$, say, yielding $2 \times 15 = 30$ potential components in total.  Jaccard similarity could be used to identify that subset of components which had the least overlap in detected genes.   In this way, a larger number of components could be obtained without use of high factorization ranks which we have seen to be unstable.
\item
{\bf Cross-dataset factorization stability}.  We went to considerable effort -- through bootstrap sampling and cluster analysis -- to select factorization ranks which we hoped would be robust and generalise well to other datasets.   A way to confirm this robustness would be to perform the factorization / clustering pipeline on \emph{both} datasets separately, yielding two sets of metagenes.  In a perfect world, these metagenes would pair up identically (in-so-far as the two datasets had identical technical characteristics, drawn from identical population of patients).  The Jaccard similarity heatmap could be used to verify this, and reject components which showed low cross-dataset similarity.
\item
{\bf Contrasting gene expression patterns between cancers}.  The current work set out to find patters of gene expression in HGSOC.  Yet it is hard to say the highlighted patterns are specific to HGSOC, or generic to all cancers.  One way of teasing out cancer specific patterns might be to take a similar approach to the previous item, but instead of looking for metagenes which are consistent between datasets, find those which are distinct.  That said, there will likely be many more sophisticated approached in the literature, given the clinical importance of the topic.
\item
{\bf Systematic comparison with published gene expression patterns}.  In analysing the meaning of each metagene above, some tentative links were made with the research literature.   But this was anecdotal and frankly not particularly scientific; searching on PubMed for GO terms associated with HGSOC and looking at one or two hits!  A more systematic and ideally automated approach is required.  The \href{https://www.ncbi.nlm.nih.gov/geoprofiles}{Geo Profiles database} at NCBI, or the \href{https://www.ebi.ac.uk/gxa/home}{Expression Atlas} at EMBL-BI might be possible starting points.

@@

\begin{figure}[htb!]
\begin{center}
\includegraphics[scale=0.7]{\plotspath{HeatMaps/clustered_heatmap_TCGA_AOCS.pdf}} 
\caption{Heatmap of the metasamples matrix for the AOCS dataset, computed from metagenes factorized from the TCGA dataset.   Several patient metadata variables are shown in columns to the left.}
\label{fig-heatmap-TCGA-AOCS}
\end{center}
\end{figure}

@@
